Understanding fine temporal scale evolutionary dynamics with lineage metrics

### 2018-04-03 paper.tex

\usepackage{natbib,alifeconf}  %% The order is important

@@

\title{Understanding fine temporal scale evolutionary dynamics with lineage metrics}
\author{Emily Dolson$^{1,2,3}$, Alexander Lalejini$^{1,2, 3}$, Steven Jorgensen$^{1,2}$ \and Charles Ofria$^{1,2, 3}$ \\
\mbox{}\\
$^1$BEACON Center for the Study of Evolution in Action, Michigan State University, East Lansing, MI, 48824 \\
$^2$Department of Computer Science and Engineering, Michigan State University, East Lansing, MI, 48824 \\
$^3$Ecology, Evolutionary Biology, and Behavior Program, Michigan State University, East Lansing, MI, 48824 \\
dolsonem@msu.edu} % email of corresponding author

@@

\begin{abstract}

@@

Evolution is an emergent effect of events (\textit{e.g.} replication, variation, and competition) that occur on a much finer-grained temporal scale. While evolution's emergent nature is a large part of why it is so fascinating, it also presents challenges to studying the short-term mechanisms that govern long-term results. In computational evolutionary systems, we theoretically have adequate data to untangle these mechanisms. In practice, though, gathering appropriate data to do so can be challenging. Moreover, because of the vast difference between the scales that evolutionary causes and effects can occur on, the quantity of such data can be overwhelming. Both of these problems can be mitigated with a standardized set of diagnostic summary metrics that paint a picture of shorter-term evolutionary dynamics within a population. Here, we suggest a suite of metrics that operate on lineages and phylogenies. We then use experimental results to provide an intuition for what they can tell us about evolution.

A lineage describes a continuous line of descent, linking ancestors and their descendants. A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation. Indeed, lineage analyses are a powerful tool for disentangling evolutionary dynamics in both natural and digital systems; digital systems, however, allow for fine-grained lineage tracking at resolutions that are impossible in modern wet lab experiments, allowing us to replay the tape of life in exacting detail and to tease apart the evolutionary recipe for whatever complex feature or phenomenon we happen to be interested in~\citep{mcphee_using_2016}. 
In a more famous example, Lenski \textit{et al.}, used the lineage of an evolved digital organism in Avida to tease apart, step-by-step how a complex feature (the capacity to perform the equals logical operation) emerged~\citep{lenski_evolutionary_2003}.

@@

%@AML:  It might be nice to wrap this section up with a little bit of a 'call to action'/'why should you care about this as a community'
In addition to demonstrating a range of metrics that are immediately applicable and useful to digital evolution research, we hope this work begins a conversation within the computational evolution community about how we quantify, interpret, and compare observed evolutionary histories. There has been extensive efforts to improve our ability to visualize both lineages and phylogenies [CITATIONS], which are incredibly valuable for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history. However, we are unaware of efforts to formalize a suite of quantitative lineage- and phylogeny-related metrics for computation evolution. Indeed, .... % people have done stuff with lineages
% - often a component of a larger experiment-specific analysis or limited to qualitative descriptions/visualizations
% - well-defined metrics not only provide valuable tools for teasing apart evolutionary dynamics, they also facilitate allow us to move away from exclusively qualitative descriptions/understandings of results ==> better quantitative understanding/statistics ==> better hypothesis testing
% - facilitate comparisons among systems. 
% - Something about activity metrics?

@@

Code for all of our metrics is open source and available in the Empirical library (https://github.com/devosoft/Empirical). Empirical is a C++ library built to facilitate writing efficient and easily sharable scientific software. To facilitate including useful features without excessive overhead, Empirical is header-only, meaning these metrics can be imported into an existing project with minimal effort.
% [QUESTION] @AML: Is it actually true that someone could include these metrics with minimal effort? How much of Empirical would it impose on the user? Just the systematics manager?

@@

Lineage length describes the number of ancestral states traversed by a lineage. If a state represents a single individual along a lineage, lineage length is a measure of the number of generations on a lineage. This form of the lineage length metric is most useful in systems where generational turnover is not algorithmically determined but instead determined by the life history strategies of organisms. In these systems, lineage length can provide a proxy for replication rate where, for lineages that span equal lengths of time, longer lineages imply faster replication rates and shorter lineages imply slower replication rates. 
%@ELD: I've always heard these called r- and k- selected rather than replicator, but I'm totally willing to believe these terms are used too.
%@AML: Yup. I have no idea if I've ever heard them referred to as 'r-replicator vs. k-replicator'. I'm going to get rid of the two parentheticals for now.

Lineage length becomes an increasingly flexible and informative metric if we begin to consider more abstract definitions of an ancestral state along a lineage. 
We might measure lineage length where an ancestral state represents a temporally continuous sequence of individuals that share a phenotypic or genotypic characteristic of interest. In these cases, lineage length only increases if the characteristic of interest changes from parent to offspring. For example, in an environment where organisms must perform a set of tasks to be successful, we might define ancestral state to be the set of tasks performed by an individual (\textit{e.g.} the set of logic operations performed by a digital organism in the default logic 9 Avida environment). In this scenario, lineage length would only increase when the set of tasks performed by an ancestor changes, compressing sequential ancestors that perform the identical sets of tasks into a single ancestral state. 
% This could be made very clear with a good example here. 
%@ELD Maybe I'm just getting tired, but I didn't quite follow that last sentence
%@AML: It was definitely hard to follow. Hopefully it's better now? 
%  -- Still could use a sentence about what the measure might tell us in the given example.

\subsubsection{Mutation Accumulation}
Mutation Accumulation defines a set of measurements that track mutational changes across a lineage. These changes can be measured in a variety of ways, such as the magnitude of the change (for real-valued genomes), or the total count of changes (for discrete-valued genomes). The type of a mutation can also be tracked, and can provide insight into the distribution of mutation types for a given lineage. Measures of mutation accumulation along the lineages of successful extant individuals can help to tease apart the relative importance of different types of mutational events when compared to what is expected by chance.

In conjunction with collected fitness information, the class of a mutation (\textit{e.g.} beneficial, deleterious, or neutral) can also be tracked. Different evolutionary conditions are expected to cause different distributions of mutations along a lineage~\citep{barrick2013genome}; deviations revealed by measures of mutation accumulation can act as a barometer for unexpected evolutionary dynamics.   
The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape as well as how well a lineage was able to cross fitness valleys~\citep{. 
Similarly, an elevated measure of neutral mutations relative to beneficial or deleterious mutations can tell us that the fitness landscape has a large neutral area that the lineage is spending most of its time drifting around.

@@

Phenotypic volatility addresses the rate at which the lineage changes phenotype (although the same concept can be applied to other types of state). In systems with discrete/categorical phenotypes, this can be measured by summing the number of times the phenotype changes. A related but subtly different measurement in such systems is the number of unique phenotypes on a lineage. In most cases, these values will be very similar. A discrepancy between them, however, would be a clear signal of interesting dynamics. Specifically, it would suggest that the lineage was in a portion of the landscape where it could rapidly switch among different phenotypes. Such behavior could be indicative of some form of evolutionary bet-hedging.

In systems with continuous-valued phenotypes, a subtly different approach is needed to measure phenotypic volatility, because there are no discrete state transitions. Instead, we can measure the overall variance in phenotype along a lineage. In some cases, it may be desirable to smooth out the noise inherent in a real-valued phenotype. We can do so by instead taking the variance of the rolling mean of fitness, to more closely approximate the idea of measuring phase transitions.
% TODO

@@

A single representative lineage can be chosen by selecting the lineage of the dominant organism (either the most fit or the more numerous; here we use the most fit). In populations with a diversity of coexisting strategies this approach can be uninformative as the dominant lineage is not necessarily representative of all successful lineages. An alternative in these scenarios is to filter out lineages that do not have offspring some predetermined number of generations later as such lineages were likely not representative of an important subset of the population. Still, any approach based on measuring only a subset of lineages can be challenging to interpret when the current dominant lineage (or lineages) is replaced with a different one; such changes can introduce a discontinuity if the value is being measured over time. If graceful responses to changes in which lineage is dominant are required, it can be advantageous to instead measure summary statistics (e.g. mean, variance, and range) across the entire population.

In scenarios where selective sweeps happen rapidly, the dominant lineage will likely be very similar to the average lineage in the population, as most of the population will be closely related. When the population contains more phylogenetic diversity, however, it is possible for the dominant lineage to differ from the mean. The nature of such differences is likely informative about the evolutionary dynamics occurring in the population. Since we have yet to observe an instance of this phenomenon, we do not explore it further in this paper. Nevertheless, it presents an argument for measuring the dominant lineage in addition to gathering summary statistics across the population.

@@

These metrics operate on entire phylogenies rather than single lineages within a population. Thus, in most cases, summarizing them across the population is not an issue as it is with lineage metrics. Because they factor in the entire population, phylogeny metrics can be more computationally expensive to calculate than metrics about a single lineage. On the other hand, because most lineages in the population share a substantial amount of history, phylogeny metrics can often be calculated more rapidly than lineage metrics can be calculated across an entire population. Note that phylogenies can be constructed with regard to any taxonomic level of organization, be it individual, genotype, phenotype, etc. Thus, when we refer to items in a phylogeny, we will use the term \textit{taxa}.

A standard technique for saving memory and time when working with lineages in computational systems is to "prune" them, removing any taxa that are not either currently alive or the ancestor of taxa that are alive. Since all of the phylogeny metrics we discuss here are borrowed from natural systems (where we do not have information about taxa without offspring), they all are designed to work on pruned phylogenies. Thus, for the remainder of this paper, we will assume we are only working with pruned phylogenies.

@@

An important distinction between phylogenies in natural vs computational systems to be aware of is that in natural systems phylogenies are generally inferred from extant taxa, whereas in computational systems they are directly measured. The inferred phylogenies used in natural systems do not contain internal nodes unless those nodes are branch points. They also do not contain history prior to the most recent common ancestor (MRCA) of all organisms in the extant population. For consistency, we will exclude pre-MRCA taxa from our analyses. However, we will not remove non-branching internal nodes, as these only serve to make our phylogenies more informative.

@@

The depth of the MRCA (i.e. how many steps from the beginning of time did it occur?) is a useful metric because of the relatively high ratio of informativeness to ease of calculation. A more recent MRCA implies more frequent selective sweeps and less long-term stable coexistence between clades. The depth of the MRCA is so easily calculated because most any implementation of phylogeny pruning requires keeping track of it already. Measuring the frequency with which the MRCA changes (i.e. the number of coalescence events) can also by informative, as some conditions can inflate the length of the lineage relative to other conditions, without actually increasing the frequency of selective sweeps. This scenario is particularly likely when the population size is changing over time. A downside to the depth of MRCA as a metric is that any population that does have a stable ecology will likely never see a change in the MRCA after the very beginning of evolution.

@@

Measurements of phylogenetic richness quantify the total amount of evolutionary history that is contained in a set of taxa. The most traditional metric of phylogenetic richness is the one simply titled "Phylogenetic Diversity", which is calculated as the number of the nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}. Another approach is to calculate the pairwise distances between all taxa and sum them~\citep{tucker_guide_2017}. A third approach is to calculate evolutionary distinctiveness, a measurement of a taxon's evolutionary uniqueness~\citep{isaac_mammals_2007}, and sum it across all extant taxa~\citep{tucker_guide_2017}.

@@

Measurements of phylogenetic divergence quantify how distinct the taxa in the population are from each other, and are often the means of values that can be calculated for individual taxa. For instance, one option is to calculate the mean of the pairwise distances between all taxa in the population~\citep{webb_exploring_2000}. Similarly, phylogenetic divergence can also be calculated by calculating the evolutionary distinctiveness of each taxon in the population (as alluded to in the previous section) and taking the average of these values.

@@

Measurements of phylogenetic regularity quantify how balanced the branches of the phylogeny are, and are often the variances of values that can be calculated for individual taxa. Just as the mean of the pairwise distances between all taxa in the population is a measurement of phylogenetic divergence, taking their variance is a measurement of phylogenetic regularity. The same is true of the variance of evolutionary distinctiveness across the population.

\section{Test Problems}
To understand the metrics defined above, the test problems used need to be well understood and studied. The benchmark functions from the GECCO Competition on Niching Methods meet both of these requirements and allow us to visualize the actual fitness landscape, due to the low dimensionality of the problems ~\citep{li_benchmark_2013}. For each problem, the X and Y coordinates offered by a given organism are translated by the function into a fitness value. We chose a diverse subset of these functions (Himmelblau, Shubert, Composition Function 2, and Six-Humped Camel Back) as our test problems in order to gain a broad understanding of our metrics. We used the implementations of these problems at https://github.com/mikeagn/CEC2013/ (C++ for fitness calculations during evolution, Python for post-hoc analysis). Figure \ref{TODO} shows the fitness landscapes defined by each of our four chosen test problems.

% @AML: It would be neat to have a tournament size = population size treatment (pure hill climbing)
For each test problem, we evolved populations of 1000 digital organisms under a range of mutation rates and selection strengths for 5000 generations. Each digital organism's genome consisted of two floating point numbers that defined its position in the fitness landscape. To determine which organisms reproduced each generation, we used tournament selection. We evolved populations under five different tournament sizes: one, two, four, eight, and sixteen.  Tournament size represents strength of selection where higher tournament sizes correspond to a higher strength of selection and lower tournament sizes correspond to a lower strength of selection. A tournament size of one is equivalent to no selection pressure (\textit{i.e.} every organism in the population has an equal chance of being selected to reproduce). Organisms selected to reproduce did so asexually. Each value in an offspring's genome was mutated by adding noise given by a normal distribution with a mean of 0; the `mutation rate' of a treatment defined the standard deviation used to defined this normal distribution and was given as a proportion of the test problem's domain.  For each problem and tournament size, we evolved populations at eight mutation rates: \textbf{TODO}. 
% @AML: Don't love how I've phrased mutation rate sentences.

@@

\section{Implementation}

@@

%To this end, we chose problems for which we could perfectly visualize the fitness landscape, and kept track of the complete evolutionary history of all members of the population. With these data, we were able to overlay successful lineages on top of the fitness landscape. Visualizing all lineages in each experimental condition in this way gave us an efficient check on our understanding of the underlying evolutionary dynamics in that condition. 
In order to make these metrics useful in the long run, it is critical that we have an accurate understanding of how various measurements correspond to the actual behavior of lineages. The most direct way to confirm our expectations is to visualize the path that each lineage takes across the fitness landscape, mapping the x, y, and z (fitness) coordinates of each ancestor of each member of the population. Creating such a visualization entails incorporating a large quantity of information into a limited space. When projected onto two dimensions, lineages can obscure parts of the fitness landscape (and each other). To mitigate this problem, we built a thee-dimesnional visualization using the A-Frame Framework [cite].

@@

\section{Results and Discussion}
% Changing mutation: here's what happens when we change mutation on each problem
\begin{figure*}
\includegraphics[width=7in]{figs/all_mutation_rate.png}
\caption{Values of example metrics across different mutation rates for each of the four problems. Circles are medians and vertical lines show inter-quartile range.}
\end{figure*}

% \begin{figure}
% \includegraphics[width=3.5in]{figs/mrca_depth_mutation.png}
% \end{figure}

% \begin{figure}
% \includegraphics[width=3.5in]{figs/variance_deleterious_mutation_rate.png}
% \end{figure}

% - Metric 1
% Problem 1
% Problem 2
% Problem 3
% Problem 4
% - Metric 2
% ...

\begin{figure*}
\includegraphics[width=7in]{figs/all_ts.png}
\caption{Values of example metrics across different tournament sizes for each of the four problems. Circles are medians and vertical lines show inter-quartile range.}
\end{figure*}
% \begin{figure}
% \includegraphics[width=3.5in]{figs/mrca_ts.png}
% \end{figure}

% Changing selection
% - Metric 1
% Problem 1
% Problem 2
% Problem 3
% Problem 4
% - Metric 2

>>>>

\usepackage{natbib,alifeconf}  %% The order is important
\usepackage{amsmath}
\usepackage{hyperref}

@@

% Alternative titles:
%  - A Suite of Metrics for Understanding Evolutionary Dynamics/Histories
\title{Understanding fine temporal scale evolutionary dynamics with lineage metrics}
\author{Emily Dolson$^{1,2,3}$, Alexander Lalejini$^{1,2, 3}$, Steven Jorgensen$^{1,2}$ \and Charles Ofria$^{1,2, 3}$ \\
\mbox{}\\
$^1$BEACON Center for the Study of Evolution in Action, Michigan State University, East Lansing, MI, 48824 \\
$^2$Department of Computer Science and Engineering, Michigan State University, East Lansing, MI, 48824 \\
$^3$Ecology, Evolutionary Biology, and Behavior Program, Michigan State University, East Lansing, MI, 48824 \\
dolsonem@msu.edu} % email of corresponding author

@@

\begin{abstract}
We suggest a suite of diagnostic metrics that operate on lineages and phylogenies in digital evolution experiments with the aim of improving our capacity to quantitatively explore evolutionary histories in digital evolution experiments.
We present three types of lineage measurements: lineage length, mutation accumulation, and phenotypic volatility. 
Additionally, we suggest the adoption of four phylogeny measurements from biology: most-recent common ancestor depth, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity. 
We demonstrate the use of each metric on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths, confirming our intuitions about what they can tell us about evolutionary dynamics.

@@

Evolution is a collective effect of many smaller events such as replication, variation, and competition that occur on a much finer-grained temporal scale. While evolution's emergent nature is a large part of why it is so fascinating, it also presents challenges to studying the short-term mechanisms that govern long-term results. In computational evolutionary systems, we can collect data to help untangle these mechanisms. In practice, though, gathering this data can be challenging. 
Moreover, because evolutionary effects can be witnessed on a vastly different scale from their causes, the quantity of such data can be overwhelming.
Both of these problems can be mitigated with a standardized set of diagnostic metrics that summarize shorter-term evolutionary dynamics within a population. Here, we suggest a suite of such metrics that operate on lineages and phylogenies. We then use experimental results to provide an intuition for what they can tell us about evolution.

A lineage describes a continuous line of descent, linking ancestors and their descendants. A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation. Indeed, lineage analyses are a powerful tool for disentangling evolutionary dynamics in both natural and digital systems; digital systems, however, allow for fine-grained lineage tracking at resolutions that are impossible in modern wet lab experiments, allowing us to replay the tape of life in exacting detail and to tease apart the evolutionary recipe for whatever complex feature or phenomenon we happen to be interested in~\citep{mcphee_using_2016}. 
In a more famous example, Lenski \textit{et al.}, used the lineage of an evolved digital organism in Avida to tease apart, step-by-step how a complex feature (the capacity to perform the equals logical operation) emerged~\citep{lenski_evolutionary_2003}. %@SPJ: 'More famous example' sounds a bit weird to me

@@

%@AML:  It might be nice to wrap this section up with a little bit of a 'call to action'/'why should you care about this as a community'
%@ELD: I am in favor.
In addition to demonstrating a range of metrics that are immediately applicable and useful to digital evolution research, we intend for this work to begin a conversation within the computational evolution community about how we quantify, interpret, and compare observed evolutionary histories. There has been extensive efforts to improve our ability to visualize both lineages and phylogenies [CITATIONS], which are incredibly valuable for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history. However, we are unaware of efforts to formalize a suite of quantitative lineage and phylogeny-related metrics for computation evolution. Indeed, .... % people have done stuff with lineages
% - often a component of a larger experiment-specific analysis or limited to qualitative descriptions/visualizations
% - well-defined metrics not only provide valuable tools for teasing apart evolutionary dynamics, they also facilitate allow us to move away from exclusively qualitative descriptions/understandings of results ==> better quantitative understanding/statistics ==> better hypothesis testing
% - facilitate comparisons among systems. 
% - Something about activity metrics?

@@

Code for all of our metrics is open source and available in the Empirical library (https://github.com/devosoft/Empirical). Empirical is a C++ library built to facilitate writing efficient and easily sharable scientific software. To facilitate including useful features without excessive overhead, Empirical is header-only, meaning these metrics can be imported into an existing project with minimal effort.
% [QUESTION] @AML: Is it actually true that someone could include these metrics with minimal effort? How much of Empirical would it impose on the user? Just the systematics manager? 
%@ELD: They should just need the systematics manager. And they might need to hook up a couple of data collection signals.

@@

Lineage length describes the number of ancestral states traversed by a lineage. If a state represents a single individual along a lineage, lineage length is a measure of the number of generations on a lineage. This form of the lineage length metric is most useful in systems where generational turnover is not algorithmically determined but instead determined by the life history strategies of organisms. In these systems, lineage length can provide a proxy for replication rate where, for lineages that span equal lengths of time, longer lineages imply faster replication rates (\textit{e.g.} r-selected lineage) and shorter lineages imply slower replication rates (\textit{e.g.} K-selected lineage). 
%@ELD: I've always heard these called r- and k- selected rather than replicator, but I'm totally willing to believe these terms are used too.
%@AML: Yup. I have no idea if I've ever heard them referred to as 'r-replicator vs. k-replicator'. I'm going to get rid of the two parentheticals for now.
%@ELD: I like the idea of putting it in that context. I was just suggesting we replace "replicator" with "selected"
%@AML: Put the parantheticals back.

Lineage length becomes an increasingly flexible and informative metric if we begin to consider more abstract definitions of an ancestral state along a lineage. 
We might measure lineage length where an ancestral state represents a temporally continuous sequence of individuals that share a phenotypic or genotypic characteristic of interest. In these cases, lineage length only increases if the characteristic of interest changes from parent to offspring. For example, in an environment where organisms must perform a set of tasks to be successful, we might define ancestral state to be the set of tasks performed by an individual (\textit{e.g.} the set of logic operations performed by a digital organism in the default logic 9 Avida environment). In this scenario, lineage length would only increase when the set of tasks performed by an ancestor changes; any sequential ancestors that perform the identical sets of tasks would be compressed into a single state in the sequence. 
% This could be made very clear with a good example here. 
%@ELD Maybe I'm just getting tired, but I didn't quite follow that last sentence
%@AML: It was definitely hard to follow. Hopefully it's better now?
%@ELD: That definitely helped. I tried tweaking it a little more (I think the term "ancestral state" was confusing me, because I'm used to it being the state of the ancestor at the root of the tree)
%  -- Still could use a sentence about what the measure might tell us in the given example.

\subsubsection{Mutation Accumulation}
Mutation accumulation defines a set of measurements that track mutational changes across a lineage. These changes can be measured in a variety of ways, such as the magnitude of the change (for real-valued genomes), or the total count of changes (for discrete-valued genomes). The type of a mutation can also be tracked, and can provide insight into the distribution of mutation types for a given lineage. Measures of mutation accumulation along the lineages of successful extant individuals can help to tease apart the relative importance of different types of mutational events when compared to what is expected by chance.

In conjunction with collected fitness information, the class of a mutation (\textit{e.g.} beneficial, deleterious, or neutral) can also be tracked. Different evolutionary conditions are expected to cause different distributions of mutations along a lineage~\citep{barrick2013genome}; deviations revealed by measures of mutation accumulation can act as a barometer for unexpected evolutionary dynamics.  
% @AML: the below points likely have many things that could be cited to support. 
The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape, as well as how well a lineage was able to cross fitness valleys~\citep{covert_experiments_2013}. 
Similarly, an elevated measure of neutral mutations relative to beneficial or deleterious mutations can tell us that the fitness landscape has a large neutral area that the lineage is spending most of its time drifting around.

@@

Phenotypic volatility addresses the rate at which the lineage changes phenotype (although the same concept can be applied to other types of state). In systems with discrete/categorical phenotypes, this can be measured by summing the number of times the phenotype changes. A related but subtly different measurement in such systems is the number of unique phenotypes on a lineage. In most cases, these values will be very similar. A discrepancy between them, however, would be a clear signal of interesting dynamics. Specifically, it would suggest that the lineage was in a portion of the landscape where it could rapidly switch among different phenotypes. Such behavior could be indicative of some form of evolutionary bet-hedging~\citep{beaumont2009experimental}. % @AML: There are a whole lot of citations for bet-hedging.. Just picked a popular Nature article on experimental evolution of bet hedging. Alternatively, could go with older paper suggesting bet-hedging as a possible strategy that exists. 
%@ELD: This one sounds good. Normally I'd say cite both, but we probably don't have the space.

In systems with continuous-valued phenotypes, a subtly different approach is needed to measure phenotypic volatility, because there are no discrete state transitions. Instead, we can measure the overall variance in phenotype along a lineage. In some cases, it may be desirable to smooth out the noise inherent in a real-valued phenotype. We can do so by instead taking the variance of the rolling mean of fitness, to more closely approximate the idea of measuring phase transitions.
% SPJ: What is meant by 'rolling mean' ?
%@ELD: Would "moving average" be more familiar?

% TODO

@@

A single representative lineage can be chosen by selecting the lineage of the dominant organism (either the most fit or the more numerous; here we use the most fit). In populations with a diversity of coexisting strategies, this approach can be uninformative as the dominant lineage is not necessarily representative of all successful lineages. An alternative in these scenarios is to filter out lineages that do not have offspring some predetermined number of generations later as such lineages were likely not representative of an important subset of the population. Still, any approach based on measuring only a subset of lineages can be challenging to interpret when the current dominant lineage (or lineages) is replaced with a different one; such changes can introduce a discontinuity if the value is being measured over time. If graceful responses to changes in which lineage is dominant are required, it can be advantageous to instead measure summary statistics (e.g. mean, variance, and range) across the entire population.

In scenarios where selective sweeps happen rapidly, the dominant lineage will likely be very similar to the average lineage in the population, as most of the population will be closely related. When the population contains more phylogenetic diversity, however, it is possible for the dominant lineage to differ from the mean. The nature of such differences is likely informative about the evolutionary dynamics occurring in the population. 
%Since we have yet to observe an instance of this phenomenon, we do not explore it further in this paper. 
% @AML: Not sure what the above sentence is saying.
%Nevertheless, it presents an argument for measuring the dominant lineage in addition to gathering summary statistics across the population. % @SPJ: I am not sure what argument this is referring to? The fact that there may or may not be a dominant lineage that differs from the mean?
%@ELD: The point was just that sometimes the dominant lineage is different from the mean and that probably matters but we're not going to talk about it. Since we're going to be short on space, I think we can just cut these lines.

@@

These metrics operate on entire phylogenies rather than single lineages within a population. Thus, in most cases, summarizing them across the population is not an issue as it is with lineage metrics. Because they factor in the entire population, phylogeny metrics can be more computationally expensive to calculate than metrics about a single lineage. On the other hand, because most lineages in the population share a substantial amount of history, phylogeny metrics can often be calculated more rapidly than lineage metrics can be calculated across an entire population. Note that phylogenies can be constructed with regard to any taxonomic level of organization, be it individual, genotype, phenotype, etc. Thus, when we refer generally to items in a phylogeny, we will use the term \textit{taxa}. % @SPJ: could reword to 'Thus, when we refer generally to items in a phylogeny, we will use the term \textit{taxa}.',  might be more clear?
%@ELD: sounds good!

A standard technique for saving memory and time when working with lineages in computational systems is to ``prune" them, removing any taxa that are not either currently alive or the ancestor of taxa that are alive. Since all of the phylogeny metrics we discuss here are borrowed from natural systems (where we do not have information about taxa without offspring), they all are designed to work on pruned phylogenies. Thus, for the remainder of this paper, we will assume we are only working with pruned phylogenies.

@@

An important distinction between phylogenies in natural versus computational systems to be aware of is that in natural systems phylogenies are generally inferred from extant taxa, whereas in computational systems they are directly measured. The inferred phylogenies used in natural systems do not contain internal nodes unless those nodes are branch points. They also do not contain history prior to the most recent common ancestor (MRCA) of all organisms in the extant population. For consistency, we will exclude pre-MRCA taxa from our analyses. However, we will not remove non-branching internal nodes, as these only serve to make our phylogenies more informative.

@@

The depth of the MRCA (i.e. how many steps from the beginning of time did it occur?) is a useful metric because of the relatively high ratio of informativeness to ease of calculation. A more recent MRCA implies more frequent selective sweeps and less long-term stable coexistence between clades. The depth of the MRCA is easily calculated because most implementations of phylogeny pruning require keeping track of it already. Measuring the frequency with which the MRCA changes (i.e. the number of coalescence events) can also be informative, as some conditions can inflate the length of the lineage relative to other conditions without actually increasing the frequency of selective sweeps. This scenario is particularly likely when the population size is changing over time. A downside to the depth of MRCA as a metric is that any population that does have a stable ecology will likely never see a change in the MRCA after the very beginning of evolution (which is informative as a way of detecting stably coexisting branches in the phylogeny). % @SPJ: change to 'The depth of the MRCA is easily calculated because most implementations of phylogeny pruning require ...' (i.e remove 'so' and 'most any')
% @SPJ: The above sentence might need use some support (why is it informative?)
%@ELD: Better?

@@

Measurements of phylogenetic richness quantify the total amount of evolutionary history that is contained in a set of taxa. The most traditional metric of phylogenetic richness is the one simply titled ``Phylogenetic Diversity", which is calculated as the number of the nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}. Another approach is to calculate the pairwise distances between all taxa and sum them~\citep{tucker_guide_2017}. A third approach is to calculate evolutionary distinctiveness, a measurement of a taxon's evolutionary uniqueness~\citep{isaac_mammals_2007}, and sum it across all extant taxa~\citep{tucker_guide_2017}.

@@

Measurements of phylogenetic divergence quantify how distinct the taxa in the population are from each other and are often the means of values that can be calculated for individual taxa. For instance, one option is to calculate the mean of the pairwise distances between all taxa in the population~\citep{webb_exploring_2000}. Similarly, phylogenetic divergence can also be calculated by calculating the evolutionary distinctiveness of each taxon in the population (as alluded to in the previous section) and taking the average of these values.

@@

Measurements of phylogenetic regularity quantify how balanced the branches of the phylogeny are and are often the variances of values that can be calculated for individual taxa. Just as the mean of the pairwise distances between all taxa in the population is a measurement of phylogenetic divergence, taking their variance is a measurement of phylogenetic regularity. The same is true of the variance of evolutionary distinctiveness across the population.

\section{Test Problems}
\begin{figure}
\includegraphics[width=3.5in]{figs/landscapes.png}
\caption{The fitness landscapes used in this experiment: A) Himmelblau, B) Six-humped Camel Back, C) Shubert, and D) Composition Function 2. Interactive versions available at \url{https://emilydolson.github.io/fitness_landscape_visualizations}.}
\label{fig:landscapes}
\end{figure}

To understand the metrics defined above, the test problems used need to be well understood and studied. The benchmark functions from the GECCO Competition on Niching Methods meet both of these requirements and allow us to visualize the actual fitness landscape, due to the low dimensionality of the problems ~\citep{li_benchmark_2013}. For each problem, the X and Y coordinates offered by a given organism are translated by the function into a fitness value. We chose a diverse subset of these functions (Himmelblau, Shubert, Composition Function 2, and Six-Humped Camel Back) as our test problems in order to gain a broad understanding of our metrics. We used the implementations of these problems at https://github.com/mikeagn/CEC2013/ (C++ for fitness calculations during evolution, Python for post-hoc analysis). Figure \ref{fig:landscapes} shows the fitness landscapes defined by each of our four chosen test problems.

% @AML: It would be neat to have a tournament size = population size treatment (pure hill climbing)
For each test problem, we evolved populations of 1000 digital organisms under a range of mutation rates and selection strengths for 5000 generations. Each digital organism's genome consisted of two floating point numbers that defined its position in the fitness landscape. 
We initialized populations by randomly generating a number of organisms equal to the population size. 
To determine which organisms reproduced each generation, we used tournament selection. We evolved populations under five different tournament sizes: one, two, four, eight, and sixteen.  Tournament size represents strength of selection where higher tournament sizes correspond to a higher strength of selection and lower tournament sizes correspond to a lower strength of selection~\citep{blickle1995mathematical}.
A tournament size of one is equivalent to no selection pressure (\textit{i.e.} every organism in the population has an equal chance of being selected to reproduce). Organisms selected to reproduce did so asexually. Each value in an offspring's genome was mutated by adding noise given by a normal distribution with a mean of 0; the `mutation rate' of a treatment defined the standard deviation used to defined this normal distribution and was given as a proportion of the test problem's domain. 
We prevented mutations from causing a value from exceeding the valid domain of the given problem. 
For each problem and tournament size, we evolved populations at eight mutation rates: $1\text{e-}08$, $1\text{e-}07$, $1\text{e-}06$, $1\text{e-}05$, $1\text{e-}04$, $1\text{e-}03$, $1\text{e-}02$, and $1\text{e-}01$. 
% [.10, 0.01, 0.001, 0.0001, 0.00001,  0.000001, 0.0000001, 0.00000001]
For all experimental conditions, we ran ten replicates, each with a unique random number seed. 
Our experiment is implemented using the Empirical library; our implementation can be found at https://github.com/stevenjson/ALife2018-Lineage. % @AML: We might want to rename the repository to match whatever the title of this paper ends up being (e.g. ALIFE2018-TITLE).

@@

@@

%To this end, we chose problems for which we could perfectly visualize the fitness landscape, and kept track of the complete evolutionary history of all members of the population. With these data, we were able to overlay successful lineages on top of the fitness landscape. Visualizing all lineages in each experimental condition in this way gave us an efficient check on our understanding of the underlying evolutionary dynamics in that condition. 
In order to make these metrics useful in the long run, it is critical that we have an accurate understanding of how various measurements correspond to the actual behavior of lineages. The most direct way to confirm our expectations is to visualize the path that each lineage takes across the fitness landscape, mapping the x, y, and z (fitness) coordinates of each ancestor of each member of the population. Creating such a visualization entails incorporating a large quantity of information into a limited space. When projected onto two dimensions, lineages can obscure parts of the fitness landscape (and each other). To mitigate this problem, we built a thee-dimensional visualization using the A-Frame Framework [cite].

@@

\section{Results and Discussion}

% Figure X and Y show results @ details... 
% All combinations can be found in supplementary material.

% Changing mutation: here's what happens when we change mutation on each problem
\begin{figure*}
\includegraphics[width=7in]{figs/all_mutation_rate.png}
\caption{Values of example metrics across different mutation rates for each of the four problems. All experiments shown here used a tournament size of 4. Circles are medians and vertical lines show inter-quartile range.}
\label{fig:mutation_rate}
\end{figure*}

Overall, our results were consistent with what evolutionary theory would predict. As mutation rate increases, coalescence takes longer, as evidenced by the fact that the MRCA is farther back in time at higher mutation rates (see Figure \ref{fig:mutation_rate}). Consequently, phylogenetic richness (as measured by phylogenetic diversity) is higher at higher mutation rates. Phylogenetic divergence, measured here as mean pairwise distance between taxa, is similarly higher at higher mutation rates. Evolutionary distinctiveness, being another measurement of phylogenetic divergence, behaved almost identically (see Supplementary Material). Variance of evolutionary distinctiveness and pairwise distance between taxa (phylogenetic regularity metrics) behaved very similarly to the phylogenetic divergence metrics. This pattern makes sense, as most phylogenetic divergence on these landscapes will produce unbalanced phylogenetic trees. If there were stable coexistence between multiple clades, we would expect to see a reduced correlation between the phylogenetic divergence metrics and the phylogenetic regularity metrics. Increased mutation rate also increases the number of deleterious steps taken, a logical consequence of increasing the amount of mutation relative to the strength of selection.

\begin{figure*}
\includegraphics[width=7in]{figs/all_ts.png}
\caption{Values of example metrics across different tournament sizes for each of the four problems. All experiments shown here used a mutation rate of .001. Circles are medians and vertical lines show inter-quartile range.}
\label{fig:ts}
\end{figure*}
